---
title: "PCA loading permutation test"
output:
  word_document:
    reference_docx: /Users/verta/RMarkdown/word_style_1.docx
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
library(parallel)
library(gtools)
```

# In this analysis, we define a false discovery rate (FDR) for parallel diverged genes based on PCA loadings.

## Approach
1. for each composite outlier gene shuffle the sample labels while keeping all other data as original
2. re-calculate PCA and composite PC
3. store the composite PC loading from the permutation
* for each composite PC outlier, perform steps 1-3 to all possible combinations of sampling two groups of 8 individuals from 16
* calculate Z-score and P-value for observing a loading higher than the permuted null
* correct for multiple testing (FDR)

Read in PCA loadings, calculate loading for composite PC and identify 1% extreme genes as in the main article.
```{r}
PC2 = read.table('.../Tyne_Litc_PC2_loadings.txt')
PC5 = read.table('.../Tyne_Litc_PC5_loadings.txt')

PC = data.frame(x=0.145*PC2$x+0.063*-1*PC5$x)
rownames(PC) = rownames(PC2)
extremePCglobal = rownames(PC)[which(PC$x<quantile(PC$x,probs=c(0.01)) | PC$x>quantile(PC$x,probs=c(0.99)))]
```

Plot sorted loadings for composite PC. Each bar represents composite PC loading for an individual gene. Dotted horizontal lines denote the cutoff values for 1% outliers. 
```{r}
barplot(sort(PC$x),ylim=c(-0.017,0.017), ylab="Composite PC loading")
abline(h=quantile(PC$x,probs=c(0.01)),lty=2)
abline(h=quantile(PC$x,probs=c(0.99)),lty=2)
```

Load counts, subset, calculate PCA and composite PC loadings for original data.
```{r}
# load
normCounts = read.table('.../xlocCountTableVarTransformed.txt')

# only pure ecotypes
sel = c(
'c172_P_532_F_lane8_0',
'c172_P_533_M_lane8_0',
'c172_P_532_F_lane3_0',
'c172_P_533_M_lane3_0',

'c169_P_432_F_0',
'c169_P_342_M_0',
'c208_P_321_F_0',
'c208_P_531_M_0',
'c209_P_341_F_0',
'c209_P_422_M_0',

'c212_P_454_F_0',
'c212_P_551_M_0',

'c214_P_524_F_0',
'c214_P_512_M_0',

'c363_P_FC08_F_0',
'c363_P_FC18_M_0',

'c358_P_FC12_F_0',
'c358_P_FC09_M_0',
'c357_P_FC14_F_0',
'c357_P_FC06_M_0',
'c353_P_FC05_F_0',
'c353_P_FC15_M_0')

river = c(rep('Tyne',10),rep('Forss',2),rep('Shiel',2),rep('Litc',8))

eco = c(

'M','F','M','F',

'F',
'M',
'F',
'M',
'M',
'F',

'M',
'F',

'M',
'F',

'M',
'F',

'F',
'M',
'F',
'M',
'M',
'F')

design = data.frame(row.names=sel,river = river,eco = eco)


# only Tyne and Litc pure strains
sel = c('c172_P_532_F_lane3_0',
'c172_P_533_M_lane3_0','c169_P_432_F_0',
'c169_P_342_M_0',
'c208_P_321_F_0',
'c208_P_531_M_0',
'c209_P_341_F_0',
'c209_P_422_M_0',
'c363_P_FC08_F_0',
'c363_P_FC18_M_0',
'c358_P_FC12_F_0',
'c358_P_FC09_M_0',
'c357_P_FC14_F_0',
'c357_P_FC06_M_0',
'c353_P_FC05_F_0',
'c353_P_FC15_M_0')

subNormCounts = normCounts[,sel]
subDesign = design[sel,]
subDesign$sex = c('Female','Male','Female','Male','Female','Male','Female','Male','Female','Male','Female','Male','Female','Male','Female','Male')

pca = prcomp(t(subNormCounts),scale=F)
compPC = data.frame(x=0.145*pca$rotation[,"PC2"]+0.063*-1*pca$rotation[,"PC5"])
```


## For each gene deemed as composite PC outlier (N=586)

Calculate all possible combinations of samples (modified from FCJ original).
```{r}
mylist<-seq(1,16,by=1);

#first create 12870 combinations of set_a (12870 ways to sample 8 fish from a set of 16)
myperm<-combinations(16, 8, v=seq(1,16,by=1), set=F, repeats.allowed=F)

#create an emptydataframe 
#(cols1-8 are set1, cols9-16 set2, 17 collapsed string of set1, 18 collapsed string of set 2, 19:34 sets recoded as "M" and "F")
mysets<-data.frame(matrix(vector(mode = 'numeric',length = 34*nrow(myperm)/2), nrow = nrow(myperm)/2, ncol = 34))
icounter<-0;

for (j in 1:length(myperm[,1])) {
	#progress report
	if((j%%500)==0) {
		print(j);
	}

	#extract seta from row of permutation table
	myperm_a<-myperm[j,];
	#assign names so we can use it in the following lookup
	names(myperm_a)<-myperm_a;
	#lookup values of setb by requireing that these are NOT in names of seta perma
	myperm_b<-mylist[-(myperm_a)];
	#assign names to set b
	names(myperm_b)<-myperm_b;
	#collapse the sets into a single string of numbers (eg 12345678)
	astring<-paste(myperm_a,sep='',collapse='');
	bstring<-paste(myperm_b,sep='',collapse='');

	#make a list of unique sets by only keeping the first time we've seen this perma/permb in the rownames.
	if(((astring %in% row.names(mysets))==F)&((bstring %in% row.names(mysets))==F)) {
		icounter<-icounter+1;
		mysets[icounter,]<-c(myperm_a,myperm_b,astring,bstring,seq(1,16,by=1));
		row.names(mysets)[icounter]<-astring;
		mysets[icounter,myperm_a+18]<-rownames(subDesign[which(subDesign$eco == "F"),]);
		mysets[icounter,myperm_b+18]<-rownames(subDesign[which(subDesign$eco == "M"),]);
	} else {
		#skip this because we have already seen it.
	}		
}

#extract columns with M and F as factors from mysets dataframe above.
permutedEco<-mysets[,19:34];
```


For a thousand random combinations

* shuffle values for each gene one at a time
* calculate PCA
* calculate composite PC

After thousand shuffles, compare the distribution of random generated composite PC loadings to the observed and calculate a P-value. (Loop for calculating the values commented out because takes ~20 h to run).
```{r}
# here you assign the random shuffling of samples to be tested - can be all possible combinations or a subsample of them (should be the same for all genes in the following loop)
# randomCombs = rownames(permutedEco)
# 
# # detect how many cores the machine has available
# numCores = detectCores()-2
# 
# # function that performs permutation and returns a P-value
# permutatePC = function(x){
#   print(paste("Permuting gene",which(extremePCglobal %in% x)))
#   ptm = proc.time()[3]
#   origLoad = compPC[x,"x"]
#   # assign a new randomized data frame based on original data
#   randCounts = subNormCounts
#   randLoadings = c()
#   for (y in 1:length(randomCombs)){
#     colSel = permutedEco[randomCombs[y],]
#     # shuffle the values of the permuted gene ("x"), values for all other genes remain unchanged
#     randCounts[x,] = subNormCounts[x,as.character(colSel[1,])]
#     # re-calculate PCA
#     pcaRand = prcomp(t(randCounts),scale=F)
#     compPCrand = data.frame(x=0.145*pcaRand$rotation[,"PC2"]+0.063*-1*pcaRand$rotation[,"PC5"])
#     randLoadings[y] = compPCrand[x,]
#   }
#   zScore = (origLoad-mean(randLoadings))/sd(randLoadings)
#   prob = 2*pnorm(abs(zScore),lower.tail = F)
#   print(paste("Time per gene",proc.time()[3] - ptm))
#   names(prob) = x
#   return(prob)
# }
# 
# # parallelize computation using maximum number of cores
# permutedPvalues = unlist(mclapply(extremePCglobal, permutatePC, mc.cores = numCores))
# 
# # save results
# save.image(".../PCAloadingPermutation.RData")

# load results
load(".../PCAloadingPermutation.RData")
```

Correct all P-values for multiple testing.
```{r}
PCfdr = p.adjust(permutedPvalues,method="fdr")
```


Get summary numbers for composite PC outliers
```{r}
# P-values for composite PC outliers - is the loading different from random loadings that we get for this gene when we destroy the sample structure (shuffle samples)
hist(permutedPvalues,breaks=100)

# FDR (Benjamini-Hochberg) (expected proportion of Type I errors among the rejected hypotheses)
hist(PCfdr,breaks=100, main="Distribution of FDR values",xlab="FDR")

# average FDR over all composite PC outliers
median(PCfdr)
```






